Real-space computation of $E$/$B$-mode maps I: Formalism, Compact Kernels, and Polarized Filaments

### 2018-07-23 articles/formalism/sections/discussion.tex
Edits to discussion.  Added reference to febecop.  Moved note about real-space implemented in that foreground paper to a footnote in the formalism section.

\section{Discussion}\label{sec:discussion}
In this work, we presented a first derivation of the real space operators on the sphere that transforms the Stokes vector to into a vector of scalars and vice versa. We also presented real space operators that directly decompose the full Stokes vector \vp{} into vector \vp{E} and \vp{B} that correspond to the respective scalar modes.  To facilitate these derivations we introduced a vector-matrix notation which allows for concise book keeping of all the standard operations involved in the analysis of (spin-2 fields) polarization maps on the sphere. This real-space analysis method trivially generalizes to maps of arbitrary spin.

These real space operators offer a spatially intuitive way of understanding the different decompositions of the Stokes vector on the sphere. We explicitly  demonstrated that all the real space operators are separable into azimuthal and radial parts. While the azimuthal part of the operators is primarily responsible for handling the spin decomposition, the radial weights determine the non-local dependence of the resulting fields on the original fields.  Only the radial part of the kernel depends on the band limit. 
The radial part of the operator kernel are roughly self-similar in the sense that the radial kernels evaluated with some band limit is related to another radial kernel evaluated with a different band limit by a simple rescaling of the function. We use this property to define a non-locality parameter $\beta_0$ as the angular distance at which the amplitude of the radial kernel falls below one percent of its maxima and it  is a function of the maximum multipole $\ell_{\rm max}$ available for the analysis. We empirically show that the non-locality parameter is approximately given by: $\beta_0 = \mathrm{min}(180^{\circ}, 180^{\circ} \frac{\ell_{0}}{\ell_{\rm max}})$ with $\ell_{0}=22$ for the operator that converts Stokes Q/U to scalars E/B and vice versa (see \fig{fig:rad_ker_decay}). An analysis in  \cite{Zaldarriaga2001a} treated real space $E/B$ operators in the flat sky.  It did not explicitly derive the radial part of the kernel, but argued on geometric grounds that it should fall with angular separation as $\beta^{-2}$.  We find that this agrees with the average behavior of radial functions after averaging oscillations and some such averaging takes place due to pixelization of the signal.  However, for precision reconstruction of $E/B$ on the sphere, the oscillations must be taken into account.

Our careful study of the real space operators show that they can be expressed either as a Green's function or as a convolving beam function.  The convolution interpretation is not a totally new concept.  It guides the discussion in \cite{Zaldarriaga2001a} and closely relates to the popular spokes and pinwheel descriptions of the $E/B$ modes.  However, the Green's function interpretation of the operators is a new one and is discussed here for the first time. These two different interpretations of the operators emerge from the expression of the kernels in terms of the forward or inverse rotation Euler angles.

@@

Modified real-space operators with compact kernels could open several alternative analysis routes in the future.  No spin-harmonic transforms are necessary as the real space operators only rely on computing the Euler angles which can be done on the fly.  The radial functions (depending on $P_{\ell}^{2}$) need be tabulated only once at some determined resolution.  Especially given the Green's function interpretation of these operators, their implementation is trivially parallelizable over a compact domain, since the $E/B$ contribution from the Stokes charges in each pixel can be evaluated independently.

@@

Finally, the toolbox of real-space operators gives us more intuition about the $E$- and $B$-mode structure of polarized gas and dust filaments in the Milky Way, an important foreground for inflationary science. We demonstrate that filaments with finite length or changing radius of curvature result in $B$-mode patterns in addition to the $E$-mode pattern already expected.  We therefore predict a characteristic $B$-mode pattern from filament sources that should be measurable in future polarization measurements.

>>>>

\section{Discussion}\label{sec:discussion}
In this work, we presented a first derivation of the real space operators on the sphere that transform the Stokes polarization vector to into a vector of scalars and vice versa.  We also presented real space operators that directly decompose the full Stokes vector \vp{} into vectors \vp{E} and \vp{B} that correspond to the respective scalar modes.  To facilitate these derivations we introduced a vector-matrix notation which allows for concise book keeping of all the standard operations involved in the analysis of  polarization maps (or any spin-2 fields) on the sphere. This real-space analysis method trivially generalizes to maps of arbitrary spin.

These real space operators offer a spatially intuitive way of understanding the different decompositions of the Stokes vector on the sphere. We explicitly  demonstrated that all the real space operators are separable into azimuthal and radial parts. While the azimuthal part of the operators is primarily responsible for handling the spin decomposition, the radial weights determine the non-local dependence of the resulting fields on the original fields.  Only the radial part of the kernel depends on the band limit. 
The radial parts of the operator kernels are roughly self-similar in the sense that the radial kernels evaluated with some band limit are related to other radial kernels (evaluated with a different band limit) by an approximate rescaling of the function. We use this property to define a non-locality parameter $\beta_0$ as the angular distance at which the amplitude of the radial kernel falls below one percent of its maxima and it  is a function of the maximum multipole $\ell_{\rm max}$ available for the analysis. We empirically show that the non-locality parameter is approximately given by: $\beta_0 = \mathrm{min}(180^{\circ}, 180^{\circ} \frac{\ell_{0}}{\ell_{\rm max}})$ with $\ell_{0}=22$ for the operator that converts Stokes Q/U to scalars E/B and vice versa (see \fig{fig:rad_ker_decay}). An analysis in  \cite{Zaldarriaga2001a} treated real space $E/B$ operators in the flat sky.  It did not explicitly derive the radial part of the kernel, but argued on geometric grounds that it should fall with angular separation as $\beta^{-2}$.  We find that this agrees with the average behavior of radial functions after averaging oscillations and note that some such averaging always takes place in practice due to pixelization of the signal.  However, for precision reconstruction of $E/B$ on the sphere, the oscillations must be taken into account.

Our careful study of the real space operators show that they can be expressed either as Green's functions or as convolving beam functions.  The convolution interpretation is not a totally new concept.  It guides the discussion in \cite{Zaldarriaga2001a} and closely relates to the popular spokes and pinwheel descriptions of the $E/B$ modes.  However, the radiation/Green's function interpretation of the operators is a new one and is discussed here for the first time in detail on the sphere. These two different interpretations of the operators emerge from the expression of the kernels in terms of the forward or inverse rotation Euler angles.  The mathematical forms of the Green's function and convolution kernels swap roles (and are conjugated) when transforming back to $Q/U$ from $E/B$.

@@

Modified real-space operators with compact kernels could open several alternative analysis routes in the future.  No spin-harmonic transforms are necessary as the real space operators only rely on computing the Euler angles which can be done on the fly.  The radial functions (depending on $P_{\ell}^{2}$) need be tabulated only once at some determined resolution.  Especially given the Green's function interpretation of these operators, their implementation is trivially parallelizable over a compact domain, since the $E/B$ contribution from the Stokes charges in each pixel can be evaluated independently.  \revisit{Alternatively, the spatially-varying convolution kernel could be applied a polarized effective beam, as implemented in \cite{2011ApJS..193....5M} in a parallel scheme.}

@@

Finally, the toolbox of real-space operators gives us more intuition about the $E$- and $B$-mode structure of polarized gas and dust filaments in the Milky Way, an important foreground for inflationary science. We demonstrate that filaments with finite length or changing radius of curvature result in $B$-mode patterns in addition to the $E$-mode pattern already expected.  We therefore predict a characteristic $B$-mode pattern from filament sources that should be observable in future polarization measurements.

### 2018-07-23 articles/formalism/sections/real_space_formalism.tex
Edits to discussion.  Added reference to febecop.  Moved note about real-space implemented in that foreground paper to a footnote in the formalism section.

\section{Real space polarization operators} \label{sec:real_space_operators}
\subsection{Evaluating scalars $E/B$ from Stokes $Q/U$}\label{sec:qu2eb}
In \sec{sec:pol-primer} we described the conventional procedure of computing the scalar fields $E/B$ from the Stokes parameters $Q/U$. 
In this section we derive the real space operators which can be used to directly evaluate the scalar fields $E$/$B$ on the sphere.  We use the vector-matrix notation introduced in \sec{sec:mat_pol_intro} to write down an operator equation relating the real space vector of scalars \vs to the Stokes polarization vector \vp{}:
%
\beqrys
\bar{S} &=& {{}_0\mathcal{Y}} \, \tilde T^{-1}  \, {{}_2\mathcal{Y}^{\ddagger}} \, \bar T  \bar{P}
= \frac{1}{2} {{}_0\mathcal{Y}} \, \tilde T^{\dagger} {{}_2\mathcal{Y}^{\ddagger}} \, \bar T \bar{P} \,,   \\
&=&  \bar O \bar{P}. \label{eq:qu2eb_op}
\eeqrys
%
The explicit form of the real space operator $\bar O$ can be derived by contracting over all the matrix operators. This procedure is explicitly worked out in the following set of equations:
%
\beqrys
\bar{O} &=& \frac{1}{2} {{}_0\mathcal{Y}}\, \tilde T^{\dagger} {{}_2\mathcal{Y}^{\ddagger}} \, \bar T \,, \\
&=& -0.5 \Delta \Omega \yzmat{e} \qutoxd \ymatc{q} \qutox   \,, \\
&=& -0.5 \Delta \Omega \begin{bmatrix} \sum ({}_{0}Y_e ~{}_{2}Y^{\dag}_q  +  {}_{0}Y_e~ {}_{-2}Y^{\dag}_q) & {\rm i}  \sum ({}_{0}Y_e ~ {}_{2}Y^{\dag}_q - {}_{0}Y_e ~{}_{-2}Y^{\dag}_q)  \\  - {\rm i} \sum  ({}_{0}Y_e ~ {}_{2}Y^{\dag}_q - {}_{0}Y_e~ {}_{-2}Y^{\dag}_q) & \sum ({}_{0}Y_e~ {}_{2}Y^{\dag}_q + {}_{0}Y_e ~{}_{-2}Y^{\dag}_q)  \end{bmatrix} \,, \label{eq:qu2eb_ker_1}
\eeqrys
%
where the symbol ${}_{0}Y_e$ is used to denote the sub-matrix ${}_{0}Y_{\hat{n}_e \times \ell m} \equiv {}_{0}Y_{\ell m}(\hat{n}_e)$, the symbol ${}_{\pm 2}Y^{\dag}_q$ is used to denote the transposed conjugated matrix ${}_{\pm 2}Y^*_{\ell m \times \hat{n}_q} \equiv {}_{\pm 2}Y^*_{\ell m}(\hat{n}_q)$ and the summation is over the multipole indices $\ell,m$. As before, we use the notation that the index $e$ denotes the location where the scalar fields are being evaluated, and the index $q$ denotes the location from which  the Stokes parameters are being accessed. Using the conjugation properties of the spin spherical harmonic functions it can be shown that the following identity holds true:
%
\beq
 \left [\sum_{\ell m} {}_{0}Y_{\ell m}(\hat{n}_e){}_{+2}Y^*_{\ell m}(\hat{n}_q)\right]^* = \sum_{\ell m} {}_{0}Y_{\ell m}(\hat{n}_e){}_{-2}Y^*_{\ell m}(\hat{n}_q) \,,
 \eeq
 %
where the terms on either side of the equation are those that appear in \eq{eq:qu2eb_ker_1}. Note that the operator $\bar{O}$ is real as one expects, since each sub-matrix in \eq{eq:qu2eb_ker_1} is formed by summing a complex number and its conjugate.

>>>>

\section{Real space polarization operators} \label{sec:real_space_operators}
\subsection{Evaluating scalars $E/B$ from Stokes $Q/U$}\label{sec:qu2eb}
In \sec{sec:pol-primer} we described the conventional procedure of computing the scalar fields $E/B$ from the Stokes parameters $Q/U$. 
In this section we derive the real space operators which can be used to directly evaluate the scalar fields $E$/$B$ on the sphere.  We use the vector-matrix notation introduced in \sec{sec:mat_pol_intro} to write down an operator equation relating the real space vector of scalars \vs to the Stokes polarization vector \vp{}:
%
\beqrys
\bar{S} &=& {{}_0\mathcal{Y}} \, \tilde T^{-1}  \, {{}_2\mathcal{Y}^{\ddagger}} \, \bar T  \bar{P}
= \frac{1}{2} {{}_0\mathcal{Y}} \, \tilde T^{\dagger} {{}_2\mathcal{Y}^{\ddagger}} \, \bar T \bar{P} \,,   \\
&=&  \bar O \bar{P}. \label{eq:qu2eb_op}
\eeqrys
%
The explicit form of the real space operator $\bar O$ can be derived by contracting over all the matrix operators.\footnote{While this manuscript was in preparation, work appeared in \cite{2018JCAP...05..059L} that implements the similar idea of real-space transformation of Stokes $Q/U$ parameters to their $E/B$ counterparts, but we treat it here in much more detail.}
This procedure is explicitly worked out in the following set of equations:
%
\beqrys
\bar{O} &=& \frac{1}{2} {{}_0\mathcal{Y}}\, \tilde T^{\dagger} {{}_2\mathcal{Y}^{\ddagger}} \, \bar T \,, \\
&=& -0.5 \Delta \Omega \yzmat{e} \qutoxd \ymatc{q} \qutox   \,, \\
&=& -0.5 \Delta \Omega \begin{bmatrix} \sum ({}_{0}Y_e ~{}_{2}Y^{\dag}_q  +  {}_{0}Y_e~ {}_{-2}Y^{\dag}_q) & {\rm i}  \sum ({}_{0}Y_e ~ {}_{2}Y^{\dag}_q - {}_{0}Y_e ~{}_{-2}Y^{\dag}_q)  \\  - {\rm i} \sum  ({}_{0}Y_e ~ {}_{2}Y^{\dag}_q - {}_{0}Y_e~ {}_{-2}Y^{\dag}_q) & \sum ({}_{0}Y_e~ {}_{2}Y^{\dag}_q + {}_{0}Y_e ~{}_{-2}Y^{\dag}_q)  \end{bmatrix} \,, \label{eq:qu2eb_ker_1}
\eeqrys
%
where the symbol ${}_{0}Y_e$ is used to denote the sub-matrix ${}_{0}Y_{\hat{n}_e \times \ell m} \equiv {}_{0}Y_{\ell m}(\hat{n}_e)$, the symbol ${}_{\pm 2}Y^{\dag}_q$ is used to denote the transposed conjugated matrix ${}_{\pm 2}Y^*_{\ell m \times \hat{n}_q} \equiv {}_{\pm 2}Y^*_{\ell m}(\hat{n}_q)$ and the summation is over the multipole indices $\ell,m$. As before, we use the notation that the index $e$ denotes the location where the scalar fields are being evaluated, and the index $q$ denotes the location from which  the Stokes parameters are being accessed. Using the conjugation properties of the spin spherical harmonic functions it can be shown that the following identity holds true:
%
\beq
 \left [\sum_{\ell m} {}_{0}Y_{\ell m}(\hat{n}_e){}_{+2}Y^*_{\ell m}(\hat{n}_q)\right]^* = \sum_{\ell m} {}_{0}Y_{\ell m}(\hat{n}_e){}_{-2}Y^*_{\ell m}(\hat{n}_q) \,,
 \eeq
 %
where the terms on either side of the equation are those that appear in \eq{eq:qu2eb_ker_1}. Note that the operator $\bar{O}$ is real as one expects, since each sub-matrix in \eq{eq:qu2eb_ker_1} is formed by summing a complex number and its conjugate.
